A versatile high-throughput assay based on 3D ring-shaped cardiac tissues generated from human induced pluripotent stem cell derived cardiomyocytes

Version 1

Abstract

Abstract

We developed a 96-well plate assay which allows fast, reproducible and high-throughput generation of 3D cardiac rings around a deformable optically transparent hydrogel (PEG) pillar of known stiffness. Human induced pluripotent stem cell-derived cardiomyocytes, mixed with normal human adult dermal fibroblasts in an optimized 3:1 ratio, self-organized to form ring-shaped cardiac constructs. Immunostaining showed that the fibroblasts form a basal layer in contact with the glass, stabilizing the muscular fiber above. Tissues started contracting around the pillar at D1 and their fractional shortening increased until D7, reaching a plateau at 25±1%, that was maintained up to 14 days. The average stress, calculated from the compaction of the central pillar during contractions, was 1.4±0.4 mN/mm2. The cardiac constructs recapitulated expected inotropic responses to calcium and various drugs (isoproterenol, verapamil) as well as the arrhythmogenic effects of dofetilide. This versatile high-throughput assay allows multiple in situ mechanical and structural read-outs.

Introduction

Cardiac tissue engineering (CTE) aims to generate in vitro cell constructs that recapitulate the intricate structural and functional properties of the native myocardium (van der Velden et al., 2022). As the heart lacks regenerative capacities, CTE emerged in the field of regenerative medicine with the objective of producing cardiac grafts that can be implanted in patients with heart failure (Madonna et al., 2019; Jabbour et al., 2021; Eschenhagen et al., 2022). In parallel, the rapid development of biomaterials and microfabrication techniques, combined with progress in pluripotent stem cell biology have enabled the generation of miniature engineered heart tissues (EHTs) used as “hearts-on-chips” or cardiac organoids models. While these EHTs recapitulate a limited number of functions in vitro, they have provided essential platforms for cardiac disease modeling (Filippo Buono et al. (2020); Cashman et al. (2016); Williams et al. (2021); Stillitano et al. (2016); Bliley et al. (2021); Goldfracht et al. (2019); Richards et al. (2020)) and drug development (Beauchamp et al. (2015); Polonchuk et al. (2017); Zhao et al. (2019); Mannhardt et al. (2020)), reducing the need for animal studies.

Different combinations of cells, biomaterials, and scaffolds have successfully generated 3D EHTs (Seguret et al. (2021); Zhuang et al. (2022)) with various geometries including spheroids (Beauchamp et al. (2015); Polonchuk et al. (2017); Giacomelli et al. (2020); Richards et al. (2020); Hofbauer et al. (2021); Lewis-Israeli et al. (2021)), cardiac strips (Hansen et al. (2010); Legant et al. (2009); Turnbull et al. (2014); Mannhardt et al. (2016); Nunes et al. (2013); Zhao et al. (2019)), circular bundles (Goldfracht et al. (2020); Tiburcy et al. (2017); Li et al. (2020)), myocardium-like sheets (Shadrin et al. (2017)) or ventricle-like chambers (MacQueen et al. (2018); Li et al. (2018); Lee et al. (2019)). The EHTs are optimally formed by mixing human-induced pluripotent stem cells (hiPSC)-derived cardiomyocytes, fibroblasts, and/or other cell types such as endothelial cells to more closely mimic the structure and cellular complexity of native myocardium (Giacomelli et al. (2020); Saini et al. (2015)). However, a major obstacle in the development of EHTs is the inevitable trade-off between the need for miniaturization to increase throughput (i.e. several organoids in each well of a 96- or 384-well plate) and the biological complexity of the tissues (Cho et al. (2022)). The first EHTs were generated in small batches, using pre-designed molds, and having a millimetric size (Boudou et al. (2012); Turnbull et al. (2014); Legant et al. (2009); Mannhardt et al. (2016)). Since then, two different avenues of development have emerged. On the one hand, to optimize throughput, more straightforward, low complexity assays such as spheroids were developed to obtain smaller tissues (Beauchamp et al., 2015; Polonchuk et al., 2017). These multicellular constructs are valuable models to study drug responses and the human heart micro-environment but they lack the geometric intricacy of a native cardiac tissue. On the other hand, the search for higher tissue complexity and maturity led to the development of chambered organoids (MacQueen et al. (2018); Li et al. (2018); Lee et al. (2019)). These constructs require the use of complex techniques such as bioprinting or the use of bioreactors, and tissues are generated one at a time, making these approaches time-consuming and impractical on a large scale. Therefore, the next step in tissue engineering needs to be directed towards high-throughput physiologically-relevant assays that are simple and straightforward for the end-user. Scaling down the tissue size to increase throughput, while controlling their geometry represents a major engineering challenge. Moreover in actual designs, many EHTs are attached on static posts that can impact tissue formation, result in isometric contraction and can interfere with optical acquisitions when made with opaque material.

Here, we report the development of a novel assay which allows fast, reproducible and high-throughput generation of 3D cardiac rings in a 96-well plate using hiPSC-derived cardiomyocytes. The design allows the formation of multiple ring-shaped cardiac tissues in a well. The EHTs form around a central pillar made from an optically-transparent and deformable polymer that enables in situ monitoring of the tissue contraction with simultaneous measurements of force generation.

Results

Design and characterization of the molds

3D steel molds were designed to obtain a structure on which cells can be directed to ring-shaped cavities during sedimentation (Figure 1A). The mold fits in the well of a 96-well plate, and consists of 21 identical conical structures with a central hole (Figure 1A left). The molds obtained from the manufacturer were imaged by scanning electron microscopy and found to correspond to the intended design (Figure 1A, right). Casts of the mold were created in PEG to check the shape of the inner cavities of the mold, but the resulting structures were too soft to maintain their shape in air, and were transparent in water and therefore could not be imaged. More rigid polydimethylsiloxane (PDMS) was cast from the mold to verify the shape of the resulting structures (Figure 1B). These showed that the interior profile of the cavity was shaped in a reversed hourglass, as expected.

Description of the mold and seeding procedure. A. Micromachined stainless steel mold used to shape the gel - Design (left) and scanning electron microscopy image (right) of the mold – Scale bar: 500 μm. B. Molded PEG gel (left - scale bar 2 mm) and zoomed view of a pillar from its PDMS replica (middle-scale bar 100 μm). Design and size of the pillar (right). C. Timeline of the seeding procedure. D. Cardiac rings in a well one day after seeding. Images stitched with ImageJ plugin - Scale bar: 1 mm. E. Representative compaction of a ring with time after seeding (from day 0 to day 14), in brightfield - Scale bars: 200 μm.

Figure 1—figure supplement 1. Measurement of the Young’s Modulus of PEG gel

Figure 1—figure supplement 2. Measurement of the efficiency of differentiation the the iPSCs into cardiomyocytes.

Figure 1—video 1. Multiple rings beating at D14 - Brightfield imaging x4 magnification.

Generation and characterization of the gels

3D gel structures were obtained by adding a drop of PEG gel solution on top of the molds, inverting these onto a glass coverslip, polymerizing the gel and attaching the resulting sample on a 96-well plate. To determine the stiffness of the gels, a sample of the aqueous PEG solution was polymerized. A capillary was placed perpendicularly to the edge of the gel and negative pressure was applied to the gel through the capillary. The deformation of the gel into the capillary was measured and used to determine the Young’s modulus of the gel (Figure 1 supplement 1). The values obtained (11.4±0.5 kPa) are within the physiological range for human cardiac tissue (Gershlak et al. (2013)).

Generation of ring-shaped cardiac tissues

The efficiency of the differentiation of iPSCs into cardiomyocytes was in average of 96.9 ± 0.8 % (Figure 1 Supplement 2).Cardiomyocytes derived from iPSCs (iPSC-CMs) with fibroblasts were seeded on the gels. After centrifugation, the cells sedimented into the ring-shaped cavities (Figure 1D), and compacted around the pillars to rapidly form ring-shaped cardiac constructs (Figure 1D-E). Initial tests with cardiomyocytes resulted in cell cluster formation and malformation of the rings (Figure 2A). To stabilize the tissues, iPSC-CMs were mixed with fibroblasts and different ratios of hiPSC-CMs to fibroblasts were tested in order to obtain the highest number of full stable cardiac rings with time. Three ratios of hiPSC-CMs:fibroblasts were tested (Figure 2B) based on previous reports (Tiburcy et al. (2017); Saini et al. (2015); Giacomelli et al. (2020); Beauchamp et al. (2020)). The 3:1 hiPSC-CMs:fibroblasts ratio was determined to be optimal and was used for all of the following experiments. Tissues started beating within 24 hours after seeding (Video 1). The rings compacted in time as shown in Figure 1E. In these optimization experiments, we obtained an average of 10.75±0.48 tissues per well which were stable over time for at least 20 days with the 3:1 hiPSC-CMs:fibroblasts ratio.

Composition and structure of the cardiac rings. A. Example brightfield images of cardiomyocytes seeded with or without fibroblasts one day after seeding - Scale bar: 200 μm. B. Number of full cardiac rings per well in time according to the hiPSC-CMs:fibroblasts ratio they contain (4:1,3:1 or 2:1). C. Confocal imaging of immunostained tissues at different heights: cardiac ring (top panel), and basal layer (bottom panel) of a tissue at 40X magnification. Vimentin, stained in green, corresponds to fibroblasts, troponin T, in red, is specific to cardiomyocytes, and DAPI is in blue. Scale bars: 100 μm. D. Picture of the immunostained contractile fibers at 63X magnification. Vimentin (green), Troponin T (red) and DAPI (blue) – Scale bar: 50μm. E. 3D reconstruction of a ring. x, y and z scale bars: 100 μm.

Figure 2—figure supplement 1. Larger scale pictures of panel C.

Figure 2—video 1.3D reconstruction of a ring

Structure and organization of the engineered cardiac tissues

The 3D organization of the tissues was assessed by 3D immunofluorescence imaging 14 days after seeding. The fibroblasts are concentrated at the base of the ring in contact with the glass (Figure 2C - bottom panel), whereas the cardiomyocytes form a compact ring above the fibroblasts around the central pillar (Figure 2C - top panel). Troponin T staining evidenced the formation of striated elongated fibers (Figure 2D), which are typical of cardiac tissue. The obtained cardiac tissues are 75.5±1.8 μm high and their shape is toric, as shown in the 3D reconstruction (Figure 2E and video 1).

Contractility analysis

Tissues started contracting less than 24 hours after seeding, which was visible through the deformation of their central pillar (Supplemental video 1). Each tissue was recorded in brightfield imaging, at a 10X magnification with a high-speed camera (Supplemental video 1). An in-house Matlab code was developed to quantify the contraction and relaxation phases of the tissues by monitoring the area of the central pillar over time (Figure 3A). The beat rate of the tissues first decreased between day 1 and day 3 and increased again to reach 0.58±0.030 Hz on day 14 (Figure 3B). The contraction stress developed by the tissues increased from day 1 until day 7 (Figure 3C). At day 7, the contractility parameters stabilized with an average stress of 1.4±0.1 mN/mm2 (Figure 3C.). The corresponding developed strain eA was of 24.84±0.92 % (Figure 3 supplement 1). Consistently, the relaxation and contraction speeds also increased during the first 7 days, then reaching a plateau (Figure 3D and E).

A. Principle of the in-house Matlab code used for contractility analysis: detection of the central pillar and monitoring of the evolution of its area in time, calculation and plot of the strain ϵA in time (ratio between the contraction amplitude in time and the maximum area of the central pillar) – Representative plots of the strain in time and its derivative in time, for a tissue at day 14. B. to E. Evolution of beating parameters through time after seeding at days 1, 3, 7, 10 and 14. The changes of all the parameters through time are significant (p<0.0001 - ANOVA for repeated measures -D1: n=57, D3: n=59 D7: n=47, D10: n=43, D14: n=36 tissues, from 3 differentiations). Beating parameters at each time point are compared to their value at day 1. B. Evolution of beat rate through time after seeding (***: p<0.002). C. Evolution of contraction stress through time after seeding (****: p<0.0001). D. Evolution of maximum contraction speed through time after seeding (****: p<0.0001). E. Evolution of maximum relaxation speed through time after seeding (****: p<0.0001).

Figure 3—figure supplement 1. Strain ϵA developed by the rings at D14.

Figure 3—video 1. Contractility of a ring at day 14.

Physiological testing on the engineered cardiac tissues

The response in contractility of the cardiac rings to an increase in the extracellular calcium concentration was assessed by sequential addition of calcium in the medium, starting from a concentration of 0.5 mM up to 3.5 mM (Figure 4A). The changes in the contraction stress and the contraction and relaxation speeds with calcium concentration were significant (p<0.0001 for each parameter, non significant for the beat rate). We observed a stabilization of the beat rate around 0.5Hz at 2mM and a positive inotropic response as the contraction amplitude increased by 600 % with a corresponding EC50 of 1.346 mM with a confidence interval (Cl) of [1.159;1.535] mM.

Physiological and drug testing on the cardiac tissues. Effect of the concentration in extracellular calcium (A.), verapamil (B.), isoproterenol (C.) or dofetilide (D.) on tissues contractility: beat rate, contraction stress and the maximum contraction and relaxation speeds. These parameters are expressed as a ratio between their value for each concentration and the value at basal state ([Ca2+]=0.5 mM for calcium test and [Drug]=0M for drug testing). An ANOVA for repeated measures was carried out for each parameter. For each test, the value of each parameter at each concentration is compared to its value at the minimal concentration of the drug (respectively [Ca2+]=0.5mM, [Verapamil]=10-9 M, [Isoproterenol]=10-9 M and [Dofetilide]=10-9 M). E. Poincaré plot for 0 nM, 1 nM, 3 nM and 7 nM of dofetilide. For each concentration of each drug, more than 20 tissues from 3 different concentrations could be analyzed. Data is presented as mean ±SEM. *: p<0.05;**: p<0.01;***: p<0.001;****: p<0.0001.

Pharmacological testing on the engineered cardiac tissues

We next evaluated the dose-response to increasing concentrations of several cardiotropic drugs. The changes in the beating parameters with verapamil concentration were found to be significant (p=0.0053 for the beat rate and p<0.0001 for the other parameters). The addition of verapamil, an L-type calcium channel inhibitor, induced a negative inotropic effect until the complete stop of contraction at 10μM, as expected (Figure 4B). The IC50 was estimated to be 0.677 μM, CI=[4.278;12.81].107 M.The dose-response to the β-adrenergic receptor agonist, isoproterenol was tested. We observed a trend towards increase of the contraction amplitude, as well as a significant increase in the contraction and relaxation speeds (Figure 4C), without change in the beat rate. The changes in the contraction stress and the contraction and relaxation speeds with isoproterenol concentration are significant (p=0.0004 for the contraction stress; p<0.0001 for the contraction and relaxation speeds). Finally, dofetilide, which is an inhibitor of the IKr current, was also tested at different concentrations on the cardiac rings. This drug has been shown to be pro-arrhythmic and to induce arrhythmic events such as Torsades-de-Pointes in patients. The changes in the beat rate, the contraction stress and the contraction and relaxation speeds with dofetilide concentration were found to be significant (p<0.0001 for all the parameters). As shown in Figure 4D, dofetilide significantly reduced the contraction stress and speeds until the complete stop of the contraction of the cardiac tissues for a concentration of 10 nM. Arrhythmic events were noted after adding dofetilide at concentrations of 3 and 7 nM. Indeed, the Poincaré diagram forthe different dofetilide concentrations (Figure 4E) shows an increase in the contraction-to-contraction intervals (equivalent to RR intervals) for some cells at 1 nM of dofetilide and an increasing number of erratic beats with increasing dofetilide concentrations.

Discussion

The combination of pluripotent stem cells technologies, advanced cardiomyocyte differentiation protocols and microfabrication methods have already resulted in the generation of engineered human cardiac muscle tissue-constructs, with different shapes and that can reproduce some cardiac functions (Seguret et al. (2021); Cho et al. (2022)). However, until now, approaching the complex geometry and structure of the human myocardium required complex techniques, which limits the number of tissues produced, thus hampering their application for high-throughput experiments. In this study, we aimed to address this challenge by developing a novel strategy to create ring-shaped miniature cardiac tissues around a central pillar that is made from an optically-transparent and deformable polymer. Our main findings are as follows: (1) this novel strategy enables the fast and reproducible formation of an important number of cardiac tissues (up to 21 per well in a 96-well format) with a limited number of hiPSC-cardiomyocytes (112,500 per well); (2) In this assay, fibroblasts are essential to generate and maintain the tissue structure, with a 3:1 hiPSC-CMs:fibroblast ratio providing with the best yield of tissue generation; (3) The self-organized ring-shaped cardiac tissues contract and deform the optically-transparent central pillar, which can be recorded and analyzed to estimate the contractile force developed by the tissue; (4) these EHTs display concordant responses to positive or negative inotropes, and to arrhythmogenic drugs, as required for drug testing applications.

The miniaturized cardiac constructs develop by self-organization of the hiPSC-derived cardiomyocytes and fibroblasts, which are guided by a specific design to acquire a ring shape. The resulting EHTs are in a convenient 96-well plate format and easy to use for a user familiar with cell biology techniques. The 96-well plate format is adaptable to many plate readers and automated platforms. In addition, the optical clarity of the gels and the absence of gel directly under the structures ensures optimal imaging conditions. We showed that the tissues can be fixed, stained by immunofluorescence and imaged in situ at high resolution. Moreover, our EHTs would also be compatible with live cell imaging experiments with fluorescent reporters, such as calcium transients studies or conduction studies by optical mapping. While imaging the fluorescent reporter, the contraction parameters could be monitored through the deformation of the pillar. The possibility to simultaneously record two parameters represents a significant asset. The current design of our EHTs consists of up to 21 rings in a single well, thus providing many replicates within a single experiment, even if some of the rings fail to form due to biological variability. This high throughput can compensate for potential intra-batch effects. Moreover, the possibility to seed several wells at the same time allows to easily study several differentiations, compensating for the variability which is inherent to the differentiation of hiPSCs. The predictable circular shape of the tissues makes them easy to recognize and track using software. Coupled with the known organization of the 21 rings for easy automated imaging, a single well can quickly generate data on many tissues that is quickly and automatically analyzed. The EHTs we described here are thus easy to use, image, manipulate and obtain readouts from.

We found that the circular geometry of these self-organized cardiac tissues induces a homogeneous distribution of the cardiomyocytes and consequently of the contraction forces around the pillar and in the tissue. Importantly, the presence of the central pillar not only allows resistance to be felt by the tissue, but the control of its stiffness facilitates the calculation of the force developed by the tissue. Our results showed that the tissues develop a fractional shortening (or contraction strain) of about 25%, which is close to the contraction of a human heart (Cheng et al. (2010)). The corresponding stress was 1.4 mN/mm2 which is in a similar range to other previously reported tissue-engineered cardiac muscle models with hiPSC-CMs in the literature (Seguret et al. (2021); Turnbull et al. (2014); Shadrin et al. (2017); Zhao et al. (2019); Ronaldson-Bouchard et al. (2018); Goldfracht et al. (2020); Li et al. (2020)). Some studies have reported higher force values (Ronaldson-Bouchard et al. (2018)) but in tissues that have a much large size than the tissues proposed here. The force generated by the adult human heart muscle, around 44 mM/mm2 fora tissue strip (Hasenfuss et al. (1991)), however remains significantly higher than the force developed by the currently available platforms. Few cardiac assays can precisely monitor the force exerted by the tissues. Indeed, cardiac cells seeded on micropillars (Rodriguez et al. (2014)) and gels containing tracking beads (Dou et al. (2021); Feyen et al. (2020)) provide information on local (subcellular) force exertion. The force measurements do not result from the concerted effort of an organized tissue composed of aligned cardiomyocytes, as proposed by ring-shaped tissues. Tissues attached on two posts (Turnbull et al. (2014); Mannhardt et al. (2016); Abilez et al. (2018)) can also provide some information on the force generated by a tissue but the force exertion is not radial as it is in the native heart and the presence of the posts can impact the contraction by affecting the force distribution in the tissue and creating potential edge effects in the vicinity of the posts, inducing the formation of v-necks (Abilez et al. (2018)). Lastly, some other ring-shaped cardiac constructs have to be transferred on posts or on a force transducer to measure their contraction force (Goldfracht et al. (2020); Tiburcy et al. (2017)). Therefore, the easy in situ monitoring of the force generated by our ring-shaped tissues with an image analysis routine represents a significant advantage, as less manipulation improves the degree of standardization. Finally, the stiffness of the central pillar can easily be adapted to model a pathological increase or decrease of the extracellular matrix stiffness.

The EHTs we developed are composed of cardiomyocytes that rely on fibroblasts for their assembly and inter-cellular adhesion. Fibroblasts have been shown to play a prominent role in the heart as they are crucial in the constitution of the supporting extracellular matrix and contribute to cardiomyocyte electrical coupling, conduction system insulation, vascular maintenance, and stresssensing (Ivey and Tallquist (2016); Baudino et al. (2006)). Other studies have highlighted the importance of a multi-cellular approach to generate EHTs (Giacomelli et al. (2020); Saini et al. (2015); Amano et al. (2016); Caspi et al. (2007)). In our system, the fibroblasts additionally play a role in providing a basement support under the cardiomyocytes, in contact with the glass support of the EHTs. Immunofluorescence imaging showed an organization of the cardiomyocytes parallel to each other, perpendicular to the direction of contraction, with an increased alignment of sarcomeres. Our EHTs mimic the composition and organization of the native myocardial tissue, even if only composed of cardiomyocytes and fibroblasts. Future experiments will determine whether other, less abundant, cell types of the native myocardium, such as endothelial or smooth muscle cells, can further improve heart tissue modeling.

The constructs showed a spontaneous beating in the day following seeding and this activity was maintained over several weeks thus allowing contractile measurements and pharmacological testing in the conditions of auto-pacing. As a proof-of-concept, we studied the effects of well-established negative and positive inotropes and showed the ability of our platform to reproduce the anticipated effects. We observed that our tissues presented a positive inotropic response to increasing extracellular calcium concentration with a corresponding EC50 which is in agreement with the values reported in the literature for engineered heart tissues made of hiPSC-CM, which are between 0.4 and 1.8mM for the EC50 (Feric and Radisic (2016); Schaaf et al. (2011); Streckfuss-Bömeke et al. (2013); Mannhardt et al. (2016); Turnbull et al. (2014); Goldfracht et al. (2020)). The negative inotropic response to increasing concentrations of verapamil of our constructs was also described in the literature and our IC50 value is also in the range of reported values in other EHTs: Turnbull et al. Turnbull et al. (2014) found IC50=0.61 μM and other papers range from 0.3 μM to 0.6 μM (Mannhardt et al. (2016); Thavandiran (2019)). The β-adrenergic stimulation of the tissues with increasing concentrations of isoproterenol-induced a trend towards increase in the contraction force and significantly increased contraction and relaxation speeds without change in the beat rate. This is in line with what has been shown by Goldfracht et al. (2020) and Zhao et al. (2019). Last, applying increasing dofetilide concentrations on our constructs showed a decrease in the contraction amplitude and contraction and relaxation speeds. This is in line with the literature as Zhao et al. Zhao et al. (2019) depicted a significant reduction of the action potential amplitude of the tissues above a concentration of dofetilide of 10 nM. By assessing the beating regularity by plotting each period as a function of the previous one, we highlighted an increasing number of erratic beats with the increase in dofetilide concentration. Other papers have also underlined the pro-arrhythmic effect of dofetilide on cardiomyocytes derived from iPSCs (Goldfracht et al. (2019); Patel et al. (2019); Altrocchi et al. (2020); Blinova et al. (2017)). Therefore, we demonstrated that our platform recapitulated the expected cardiac responses to several inotropes and arrhythmogenic drugs.

While representing a significant improvement, our platform presents some limitations. First, the size of the rings is very small compared to native tissue (400 μm). This size limitation is necessary for the purposes of high throughput screening, but for applications requiring larger tissues, wider rings can be considered in the future. Second, all measurements were performed under spontaneous pacing as a specific technique should be developed to allow external pacing of the tissues in the 96-well format. This includes electrical field stimulation or an optogenetics system with light stimulation. The precise and regular organization of the rings in the well makes this design very amenable to the introduction of microelectrodes for pacing, which could also allow the measurement of the action potentials of individual micro-tissues. Nevertheless, our main results show that pharmacological responses, including the detection of the pro-arrhythmogenic effects of dofetilide, were correctly measured under spontaneous beating conditions. Third, the miniaturized format of our platform creates some limitations and challenges regarding some ex-periments, including tissue stretching. Therefore, the length-tension relationships (Franck-Starling mechanism) were not determined in our tissues.

Overall the novel platform we describe here is a versatile, easy-to-use and high-throughput tool to generate cardiac rings in a reproducible way and analyze them in situ. This platform will be useful in several cardiac research fields, including disease modeling and pharmacological testing.

Methods and Materials

Mold design and characterization

Micromachined stainless steel molds were obtained from high resolution CNC machining according to our CAD designs. The designs consisted of a mold that would be slightly smaller than the footprint of a well of a 96-well plate, composed of 21 structures (see Figure 1A). To check the shape of the interior cavities of the mold, a cast of the mold was created out of polydimethylsiloxane (PDMS) (Sylgard 184, Samaro), according to manufacturer’s instructions.

3D hydrogel substrate preparation

A solution of 5% w/w photopolymerizable polyethylene glycol (PEG) was prepared in distilled water. Circular coverslips (16 mm, Paul Marienfeld GmbH, Germany) were silanized to ensure adequate bonding between the PEG and glass. A drop of 10-15 μL of gel was placed on the stainless steel mold, which was then placed in contact with a silanized coverslip. The gel solution was briefly exposed to light of the appropriate wavelength to polymerize the hydrogel. The resulting 3D structured hydrogel, bound to the coverslip, was unmolded and immediately placed in distilled water. The hydrogel was rinsed at least three times with distilled water and incubated in water overnight. For cell experiments, the coverslip coated in a 3D hydrogel was briefly dried and attached to a commercial black polystyrene 96-well plate with an adhesive backing (Grace Bio-labs, USA). The resulting well was filled with distilled water. Prior cell seeding, the plate was sterilized by UV irradiation for at least 20 minutes.

Young’s modulus measurements of the molds

The PEG-based hydrogel solutions were prepared and 100μL of solution was deposited into a ringshaped mold with an inner diameter of 1cm positioned on a 16mm glass coverslip. The solution was polymerized using 365nm UV light for 1 minute at a power of 70 mW/cm2, after which the mold was removed. The coverslip containing the gel was then transferred to the microscope (Leica camera AG, Germany). A hollow glass capillary with an inner diameter 0.75mm (World Precision Instruments, USA) was attached to a platform movable in the xy-plane which was used to position the capillary right next to the edge of the circular polymerized hydrogel. Once the capillary and the hydrogel made contact, a negative pressure was applied, using the Cobalt autonomous pressure pump (Elvesys, France), creating a small convexity in the surface of the hydrogel. Images of this convexity were taken (5X magnification) for a pressure ranging from −100mbar to −200mbar, with 10 mbar increments (Figure 1A.). Two measurements were done on two separate gels, resulting in 4 measurements per gel with 10 datapoints each. The images (and thus the size of the convexities) were analyzed using ImageJ (version 1.53c). The Youngs modulus was calculated with the formula according to literature (Gandin et al. (2021)).

Cardiomyocyte differentiation and culture

The SKiPSC-31.3 hiPSC cell line, from a healthy control subject, has been used here and was previously reported (Galende et al. (2010)). Induced pluripotent stem cells were seeded on Matrigel and cultivated in mTESR Plus medium (Stemcell Technologies). At 80 % confluency, cells were passaged with ReleSR reagent (StemCell Technologies). When hiPSCs reached a 90 % confluency, differentiation was carried out according to a protocol adapted from Garg et al. (2018). Briefly, on day 0 of differentiation, mTESR Plus was changed to RPMI-1640 (Life Technologies) supplemented with B27 without insulin (RPMI-B27 minus insulin) and 6 μM CHIR-99021 (Abcam). After 48 hours, the medium was changed to RPMI-B27 minus insulin and then to RPMI-B27 minus insulin with 5μM IWR-1 (Sigma) for 48 hours. On day 5, the medium was replaced by RPMI-B27 minus insulin, and switched to RPMI-B27 with insulin on day 7. At day 11 a glucose starvation was carried out by replacing the medium with RPMI 1640 without glucose (Life Technologies), supplemented by B27 with insulin for 3 days. Cells were then dissociated with 0.05% trypsin (Life Technologies) and replated in a 12-well plate at a density of 0.3 106cells/cm2 in RPMI-B27 with insulin. The following day, the medium was switched back to RPMI 1640 without glucose (Life Technologies), supplemented by B27 with insulin for 3 days again. These two rounds of glucose starvation have been shown to increase dramatically the percentage of cardiomyocytes obtained Sharma et al. (2015). After day 18, the cells were cultured in RPMI-B27 with insulin and the medium was changed every two days.

Cardiomyocyte characterization

The efficiency of the differentiation was assessed by cardiac troponin T flow cytometry at D21. On day 21 of differentiation, hiPSC-CMs were dissociated by enzymatic digestion (Miltenyi Multi Tissue dissociation kit 3) and stained with the Zombie NIR™ Fixable Viability Kit (BioLegend). Then, collected cell pellets were fixed and permeabilized using Inside Stain kit (Miltenyi Biotech, 130-090-477) at room temperature for 10 minutes. Cells were incubated with either APC anti-cardiac troponin T (CTNT) antibody (Miltenyi Biotech; 130-106-689 1:100) or APC isotype control (Miltenyi Biotech, 130-104-615 1:100) for 10 min at room temperature. Cells were analyzed using the BD Biosciences FACS LSR Fortessa X-20 instrument with at least 30,000 cells. Results were processed using FlowJo v10 (FlowJo, LLC).

Human fibroblasts culture

Commercially available normal adult human dermal fibroblasts (NHDF-Ad Lonza CC-2511, Lot 545147) were used up to passage 6 to generate the tissues. They were cultured in T75 flasks, in DMEM high glucose, supplemented with 10%FBS, 1% NEAA and 1% penicillin-streptomycin.

Tissues generation and culture

The wells of the assay were first rinsed twice with PBS and incubated at 37°C in DMEM high glucose, 2.3 mM CaCl2,10%FBS, 0.1% penicillin-streptomycin. To generate the tissues, cardiomyocytes were dissociated by enzymatic digestion (Miltenyi Multi Tissue dissociation kit 3) on day 22 after differentiation. Fibroblasts were dissociated with TrypLE Express Enzyme (12605010 Thermofischer). Three ratios of hiPSC-CMs:fibroblasts were tested: 4:1, 3:1 and 2:1. For the following experiments the tissues are generated with a ratio of 3:1 hiPSC-CMs:fibroblasts as this ratio allowed the best preservation of tissues in time. A total of 150k cells were seeded in each well. The required numbers of fibroblasts and cardiomyocytes were then mixed and resuspended in 200μL per well of DMEM high glucose, 2.3 mM CaCl2, 10%FBS, 0.1% penicillin-streptomycin. The plate was then incubated at 37°C for 5 minutes and centrifuged 3 times in “short” mode to ensure that the cells fall into the circular molds. The plate was then kept at 37°C and the medium was changed every two days.

Tissues recording and analysis

The cardiac rings were recorded for 10 seconds in brightfield imaging with a high speed CCD camera (PL-D672MU, Pixelink) mounted on a microscope (Primovert, Zeiss), at a 10X magnification. The videos were then analyzed with a custom Matlab script. Briefly, the area of the central pillar is monitored in time and allows to recover the contraction amplitude in time Amplitude(t)= Max(Area)-Area(t) and therefore the contraction strain in time (in terms of area) ϵA calculated as ϵA(t) = (Max(Area)-Area(t))/Max(Area) as well as the derivative of this strain in time. εA(t) is linked to the linear strain ε with the formula . As we measured the Young’s Modulus of the gel, we can recover the stress developed by the cardiac rings σ = E.ϵ(t). From these traces we also derive the contraction frequency by Fourier transform. It was considered that below a contraction of 0.5 %, the signal to noise ratio was not sufficient enough to detect the contraction of the tissue and the beating frequency was therefore set to 0 Hz. Finally, the maximum contraction and relaxation speeds of the tissue were also calculated respectively from the maxima and minima of the derivative of this strain in time.

Response to increasing extracellular calcium concentration and drug tests

Tissues response to different drugs and increasing calcium concentrations was assessed after day 14. For the increasing extracellular calcium concentrations, tissues were first changed to Tyrode solution 0.5 mM in calcium (NaCl 140 mM, KCl 5 mM, 10 mM Hepes, Glucose 10 mM, MgCl2 1 mM, CaCl2 0.5 mM, pH 7.4) and let to equilibrate for 30 minutes at 37°C, 5 % CO2. Calcium concentration was then increased by sequential addition of a Tyrode solution at 10mM in calcium. After each addition, the tissues were incubated for 5 minutes before recording. For drug tests, tissues were changed to Tyrode solution at 1.8mM in calcium (NaCl 14 mM, KCl 5 mM, 10 mM Hepes, Glucose 10 mM, MgCl2 1 mM, CaCl2 1.8 mM, pH 7.4) and let to equilibrate for 30 minutes at 37°C, 5% CO2. To evaluate the dose-response of the cardiac rings to the different drugs (Verapamil-sigma v0100000, isoproterenol - sigma 1351005, dofetilide - sigma PZ0016), the contractility was first recorded for 30 seconds in basal state and the drug was then added sequentially. Between each addition of drug, the tissues were incubated for 5 minutes at 37°C before being recorded for 30 seconds.

Immunofluorescence staining and confocal microscopy

Tissues were fixed at day 15 with 4% paraformaldehyde for 15 minutes and rinsed 3 times in PBS for 5 minutes. They were then blocked and permeabilized with blocking solution(BSA 2%, Triton 0.5%) at 4°C overnight. Primary antibodies Vimentin diluted at 1:250 (MA5-11883, ThermoFischer Scientific) and Troponin T diluted at 1:500 (ab45932, Abcam) were added the next day and incubated at 4°C overnight. Tissues were rinsed 3 times in PBS for 5 minutes and secondary antibodies (Alexa Fluor secondary antibodies, LifeTechnologies, at 1:1000) and DAPI diluted at 1:1000 were added and incubated at 4°C overnight. Immunostaining pictures were taken with a Leica SP8 confocal system.

Statistical analysis

All numerical results are expressed as mean ± standard error mean (SEM) of three independent experiments. Differences between experimental groups were analyzed with the appropriate statistical tests, specified each time. P values < 0.05 were considered significant for all statistical tests. Statistical analyses were performed with GraphPad Prism software.

eLife Assessment

The manuscript provides valuable findings that will have practical implications for high throughput cultivation of cardiac tissues. The results demonstrate convincing evidence regarding the ability to create functional cardiac tissues and maintain them over periods of days to weeks. Improving the success rate of tissue integrity over longer periods of time, reproducing the physiological isoproterenol response and studying the impulse propagation in the tissues would garner a further broad interest for the platform.

Reviewer #1 (Public Review):

The manuscript, "A versatile high-throughput assay based on 3D ring-shaped cardiac tissues generated from human induced pluripotent stem cell-derived cardiomyocytes" developed a unique culture platform with PEG hydrogel that facilitates the in-situ measurement of contractile dynamics of the engineered cardiac rings. The authors optimized the tissue seeding conditions, demonstrated tissue morphology with expressions of cardiac and fibroblast markers, mathematically modeled the equation to derive contractile forces and other parameters based on imaging analysis, and ended by testing several compounds with known cardiac responses.

To strengthen the paper, the following comments should be considered:

1. This paper provided an intriguing platform that creates miniature cardiac rings with merely thousands of CMs per tissue in a 96-well plate format. The shape of the ring and the squeezing motion can recapitulate the contraction of the cardiac chamber to a certain degree. However, Thavandiran et al (PNAS 2013) created a larger version of the cardiac ring and found the electrical propagation revealed spontaneous infinite loop-like cycles of activation propagation traversing the ring. This model was used to mimic a reentrant wave during arrhythmia. Therefore, it presents great concerns if a large number of cardiac tissues experience arrhythmia by geometry-induced re-entry current and cannot be used as a healthy tissue model. It would be interesting to see the impulse propagation/calcium transient on these miniature cardiac rings and evaluate the % of arrhythmia occurrence.

2. The platform can produce 21 cardiac rings per well in 96-well plates. The throughput has been the highest among competing platforms. The resulting tissues have good sarcomere striation due to the strain from the pillars. Now the emerging questions are culture longevity and reproducibility among tissues. According to Figure 1E, there was uneven ring formation around the pillar, which leads to the tissue thinning and breaking off. There is only 50% survival after 20 days of culture in the optimized seeding group. Is there any way to improve it? The tissues had two compartments, cardiac and fibroblast-rich regions, where fibroblasts are responsible for maintaining the attachment to the glass slides. Do the cardiac rings detach from the glass slides and roll up? The SD of the force measurement is a quarter of the value, which is not ideal with such a high replicate number. As the platform utilizes imaging analysis to derive contractile dynamics, calibration should be done based on the angle and the distance of the camera lens to the individual tissues to reduce the error. On the other hand, how reproducible of the pillars? It is highly recommended to mechanically evaluate the consistency of the hydrogel-based pillars across different wells and within the wells to understand the variance.

3. Does the platform allow the observation of non-synchronized beating when testing with compounds? This can be extremely important as the intended applications of this platform are drug testing and cardiac disease modeling. The author should elaborate on the method in the manuscript and explain the obtained results in detail.

4. The results of drug testing are interesting. Isoproterenol is typically causing positive chronotropic and positive inotropic responses, where inotropic responses are difficult to obtain due to low tissue maturity. It is inconsistent with other reported results that cardiac rings do not exhibit increased beating frequency, but slightly increased forces only. Zhao et al were using electrical pacing at a defined rate during force measurement, whereas the ring constructs are not.

Overall, the manuscript is well written and the designed platform presented the unique advantages of high throughput cardiac tissue culture. Besides the contractile dynamics and IHC images, the paper lacks other cardiac functional evaluations, such as calcium handling, impulse propagation, and/or electrophysiology. The culture reproducibility (high SD) and longevity (<20 days) still remain unsolved.

Reviewer #2 (Public Review):

The authors should be commended for developing a high throughput platform for the formation and study of human cardiac tissues, and for discussing its potential, advantages and limitations. The study is addressing some of the key needs in the use of engineered cardiac tissues for pharmacological studies: ease of use, reproducible preparation of tissues, and high throughput.

There are also some areas where the manuscript should be improved. The design of the platform and the experimental design should be described in more detail.

It would be of interest to comprehensively document the progression of tissue formation. To this end, it would be helpful to show the changes in tissue structure through a series of images that would correspond to the progression of contractile properties shown in Figure 3.

The very interesting tissue morphology (separation into the two regions) that was observed in this study is inviting more discussion.

Finally, the reader would benefit from more specific comparisons of the contractile function of cardiac tissues measured in this study with data reported for other cardiac tissue models.

Author Response

The following is the authors’ response to the original reviews.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) The author should evaluate the possibility of naturally occurring arrhythmia due to the geometry of the tissues, by using voltage or calcium dye.

Answer: We thank the reviewer for this suggestion. We have performed new experiments using a voltage-sensitive fluorescent dye (i.e. FluoVolt) with data reported in the new Figure 4 + new results section “arrhythmia analysis”. Briefly, we found that our ring-shaped tissues are compatible with live fluorescence imaging. We were then able to show that our cardiac tissues beat regularly, without naturally occurring arrhythmias or extra beats. We could not detect any re-entrant waves in our tissues in the conditions offered by the speed of our camera. A specific paragraph has also been added to the discussion.

(2) There is only 50% survival after 20 days of culture in the optimized seeding group. Is there any way to improve it? The tissues had two compartments, cardiac and fibroblast-rich regions, where fibroblasts are responsible for maintaining the attachment to the glass slides. Do the cardiac rings detach from the glass slides and roll up? The SD of the force measurement is a quarter of the value, which is not ideal with such a high replicate number.

Answer: This paper report seminal data that will serve as a foundation for further use of the platform. We are currently expanding to other cell lines with improvement in survival (see https://insight.jci.org/articles/view/161356). We confirm that the rings do not detach. The pillar was specifically designed to avoid this (See figure 1B).

As the platform utilizes imaging analysis to derive contractile dynamics, calibration should be done based on the angle and the distance of the camera lens to the individual tissues to reduce the error. On the other hand, how reproducible of the pillars? It is highly recommended to mechanically evaluate the consistency of the hydrogel-based pillars across different wells and within the wells to understand the variance.

Answer: We propose a system and a measurement method that do not need calibration. Contraction amplitude is expressed as a ratio between the contracted / relaxed areas (See figure 3 A). There is thus no influence of the distance of the camera lens.

In order to evaluate the consistency of the mechanical properties of the hydrogel, we reproduced the experiment pictured in Figure1-Supplement 1, and measured the Young’s Modulus of three different gel solutions on different days. In the three experiments performed, we found values of 10.0-12.2 kPa, resulting in a final average value of 11.2 (+/- 0.6) kPa, coherent with the value reported in the article. We are therefore confident that the mechanical properties are consistent across and within wells. More extensive mechanical characterization of the molded gels would require the access to an Atomic Force Microscope (AFM), and is considered in the future.

The author should address the longevity and reproducibility issues, by working on the calibration of camera lens position/distance to tissues and further optimizing the seeding conditions with hydrogels such as collagen or fibrin, and/or making sure the PEG gels have high reproducibility and consistency.

Answer: This paper report seminal data that will serve as a foundation for further use of the platform. This platform (including the design, approach and choice of polymers) allows a fast and reproducible formation of an important number of cardiac tissues (up to 21 per well in a 96-well format, meaning a potential total of about 2,000 tissues) with a limited number of cells.

(3) The evaluation of the arrhythmia should be more extensively explained and demonstrated.

Answer : See answer to comment 1

(4) The results of isoproterenol should be checked as non-paced tissues should have increased beating frequency with increasing dosages. Dofetilide does not typically have a negative inotropic effect on the tissues. Please check on the cell viability before and after dosing

Answer : We agree with this reviewer on the principle. However, we have repeated the experiments and we confirm our results, i.e. increasing concentrations of isoproterenol induced a trend towards increase in the contraction force and significantly increased contraction and relaxation speeds without change in the beat rate (Figure 5C). We do not have a definitive explanation for this observation. Our hypothesis is that this increase in contraction and relaxation speeds induced by isoproterenol is translated, on average in our study, into an increase in contractile force rather than in an increase in contraction frequency. This may depend on the cell line used, and is very well illustrated in a recent paper from Mannhardt and colleagues (Stem cell reports. 2020; 15(4):983–998). Of the 10 different cell lines tested in engineered heart tissues, all show an increase in contraction and relaxation speeds after isoproterenol administration, but this is translated either into an increase in contractile force (4 cell lines) or into a shortening of the beat (3 cell lines), and only 2 cell lines show an increase in both parameters. Indeed, since iPSC-CMs are immature cardiac cells, it is rare to obtain a positive force-frequency relationship without any maturation medium or mechanical or electrical training. We agree that above a concentration of 10nM, dofetilide shows cardiotoxicity in our tissues as tissues completely stop beating.

Reviewer #2 (Recommendations For The Authors):

In addition to the general comments in the public review, I have the following specific suggestions to the authors, that would help improve the manuscript.

(1) Please describe the protocol for preparation of cardiac rings (shown in Figure 1C) in more detail. In particular, please describe how the tissues were transferred from the mold into the 96-well plate and how are they positioned and characterized during the study.

Answer: There is no transfer of the tissues as they directly form in the well, that is pre-equipped with the molded PEG gel (See Figure 1B and methods section). The in situ analysis is a strong asset of this platform.

(2) Please clarify the timepoints in this study. The overall schematic in Figure 1 C shows that the rings were formed on day 22 and then studied for 14 days, while Figure 2B shows data over 20 days following seeding, and Figure 3 shows data 14 days after seeding. It appears that these were separate studies (optimization of myocyte/fibroblast ratio followed by the main study.

Answer: Figure 1C is showing the timeline including the cardiomyocytes differentiation. hiPSC-CMs are indeed seeded in the wells 22 days after starting the differentiation, which represent the Day0 for tissue formation. We apologize for the confusion.

(3) Please explain if the number of rings per well (Figure 2) was used as the only criterion for selecting the myocyte/fibroblast ratio, and if so, why. Were these rings also characterized for their structural and contractile properties?

Answer: Figure 2 supplement 1 report the contractility data according to the different tested ratios, and show no differences. The number for generated ring-shaped tissues was indeed the only criterion retained.

(4) Please provide rationale for using the dermal rather than cardiac fibroblasts.

Answer: We had previous experience generating EHTs using dermal fibroblasts which are easier to obtain commercially. Our approach could in theory also work using cardiac fibroblasts, which we have not tested in the present study.

(5) Figure 2 panels C-E show an interesting segregation of cardiomyocytes into a thin cylindrical layer that does not appear to contain fibroblasts and a shorter and thicker cylinder containing fibroblasts mixed with occasional myocytes. Please specify at which time point this structure forms, and how does it change over time in culture? At which time point were the images taken? It would be helpful to include serial images taken over 1-14 days of study.

Answer: We thank the reviewer for this interesting comment. We have performed additional immunostainings (reported in Figure 2 supplement 3) on tissues at Day 1 and day 7 after seeding. The segregation appears in the 7 first days. It appears that 1 day after seeding the fibroblasts are not yet attached, although the cardiac fiber has already started to be formed. Seven days after seeding, fibroblasts are fully spread and attached, and the contractile ring is formed and well-aligned. Brightfield images are reported in Figure 1E.

(6) In the cardiomyocyte region (Figure 2D) the cells staining for troponin seem to be only at the surfaces. The thickness of the layer is only about 30-40 µµ, so one would assume that cell viability was not an issue. Please specify and discuss the composition of this region.

Answer: We agree but we think this is a technical issue as at the center of the tissue, tissue thickness will limit laser penetration, although at the surface (inner our outer), the laser infiltrates easily between the tissue and the PEG. Moreover, we see on the zoomed view of the tissue in Figure 2 Supplement 2 that we have a staining inside the cardiac fiber, which just appears less strong due to tissue thickness.

(7) Please also discuss segregation in terms of possible causes and the implications of apparently very limited contact between the two cell types, i.e., how representative is this two-region morphology of native heart tissue. Also, it would be interesting to know how the segregation has changed with the change in myocyte/fibroblast ratio.

Answer: We are not sure there is a very limited contact as the use of fibroblasts is critical to ensure the formation of tissues (i.e. no tissues can be formed if we avoid the use of fibroblasts). We agree that these ring-shaped cardiac tissues are not especially representative of a native heart tissue in terms of interactions between several cell types. They were developed as a surrogate for physiopathological and pharmacological experiments (see a recent application in https://insight.jci.org/articles/view/161356)

(8) There is interest and demonstrated ability to culture engineered cardiac tissues over longer periods of time. Please comment what was the rationale for selecting 14-day culture and if the system allows longer culture durations.

Answer: In line with this comment, we have studied the contractile parameters of our rings 28 days after seeding and compared to their contractile parameters at D14. We found a slight increase for all the parameters, which is significant for the maximum contraction speed. Nevertheless, the data is much more variable and the number of tissues is lower (29 for D14 against 17 for D28). Therefore, we demonstrated that long-term culture of our tissues is possible, however not yet optimized. Hence, the following physiological and pharmacological tests have been done at D14.

(9) Figure 3 documents the development of contractile parameters over 14 days of culture. Would it be possible to replace the arbitrary units with the actual values? Also, would it be possible to include the corresponding images of the rings taken at the same time points, to show the associated changes in ring morphologies.

Answer: Contraction amplitude is expressed as a ratio between the contracted / relaxed areas (See figure 3 A): it is a ratio, thus without unit. Corresponding images can be seen in Figure 1 E.

(10) The measured contraction stress, strain, and the speeds of contraction and relaxation improve from day 1 to day 7 and then plateau (Figure 3, Supplemental Figure 3. Please discuss this result.

Answer: The new immunostainings performed on tissues at Day 1 and Day 7 show the progressive alignment of the cardiomyocytes and the muscular fibers, with an almost complete organization at Day 7.

(11) The beating frequency does not appear to markedly change over time, while Figure 3B shows strong statistical significance (***) throughout the 14-day period. Please check/confirm.

Answer: We confirm this result.

(12) Please comment on the lack of effect of isoproterenol on beating frequency.

Answer: We agree with this reviewer on the principle. However, we have repeated the experiments and we confirm our results, i.e. increasing concentrations of isoproterenol induced a trend towards increase in the contraction force and significantly increased contraction and relaxation speeds without change in the beat rate (Figure 5C). We do not have a definitive explanation for this observation. Our hypothesis is that this increase in contraction and relaxation speeds induced by isoproterenol is translated, on average in our study, into an increase in contractile force rather than in an increase in contraction frequency. This may depend on the cell line used, and is very well illustrated in a recent paper from Mannhardt and colleagues (Stem cell reports. 2020; 15(4):983–998). Of the 10 different cell lines tested in engineered heart tissues, all show an increase in contraction and relaxation speeds after isoproterenol administration, but this is translated either into an increase in contractile force (4 cell lines) or into a shortening of the beat (3 cell lines), and only 2 cell lines show an increase in both parameters. Indeed, since iPSC-CMs are immature cardiac cells, it is rare to obtain a positive force-frequency relationship without any maturation medium or mechanical or electrical training.

(13) Please compare the contractile function of cardiac tissues measured in this study with data reported for other iPSC-derived tissue models.

Answer : A specific paragraph tackles this aspect in the discussion

Version 2

Abstract

Abstract

We developed a 96-well plate assay which allows fast, reproducible and high-throughput generation of 3D cardiac rings around a deformable optically transparent hydrogel (PEG) pillar of known stiffness. Human induced pluripotent stem cell-derived cardiomyocytes, mixed with normal human adult dermal fibroblasts in an optimized 3:1 ratio, self-organized to form ring-shaped cardiac constructs. Immunostaining showed that the fibroblasts form a basal layer in contact with the glass, stabilizing the muscular fiber above. Tissues started contracting around the pillar at D1 and their fractional shortening increased until D7, reaching a plateau at 25±1%, that was maintained up to 14 days. The average stress, calculated from the compaction of the central pillar during contractions, was 1.4±0.4 mN/mm2. The cardiac constructs recapitulated expected inotropic responses to calcium and various drugs (isoproterenol, verapamil) as well as the arrhythmogenic effects of dofetilide. This versatile high-throughput assay allows multiple in situ mechanical and structural read-outs.

Introduction

Cardiac tissue engineering (CTE) aims to generate in vitro cell constructs that recapitulate the intricate structural and functional properties of the native myocardium (van der Velden et al., 2022). As the heart lacks regenerative capacities, CTE emerged in the field of regenerative medicine with the objective of producing cardiac grafts that can be implanted in patients with heart failure (Madonna et al., 2019; Jabbour et al., 2021; Eschenhagen et al., 2022). In parallel, the rapid development of biomaterials and microfabrication techniques, combined with progress in pluripotent stem cell biology have enabled the generation of miniature engineered heart tissues (EHTs) used as “hearts-on-chips” or cardiac organoids models. While these EHTs recapitulate a limited number of functions in vitro, they have provided essential platforms for cardiac disease modeling (Filippo Buono et al. (2020); Cashman et al. (2016); Williams et al. (2021); Stillitano et al. (2016); Bliley et al. (2021); Goldfracht et al. (2019); Richards et al. (2020)) and drug development (Beauchamp et al. (2015); Polonchuk et al. (2017); Zhao et al. (2019); Mannhardt et al. (2020)), reducing the need for animal studies.

Different combinations of cells, biomaterials, and scaffolds have successfully generated 3D EHTs (Seguret et al. (2021); Zhuang et al. (2022)) with various geometries including spheroids (Beauchamp et al. (2015); Polonchuk et al. (2017); Giacomelli et al. (2020); Richards et al. (2020); Hofbauer et al. (2021); Lewis-Israeli et al. (2021)), cardiac strips (Hansen et al. (2010); Legant et al. (2009); Turnbull et al. (2014); Mannhardt et al. (2016); Nunes et al. (2013); Zhao et al. (2019)), circular bundles (Goldfracht et al. (2020); Tiburcy et al. (2017); Li et al. (2020)), myocardium-like sheets (Shadrin et al. (2017)) or ventricle-like chambers (MacQueen et al. (2018); Li et al. (2018); Lee et al. (2019)). The EHTs are optimally formed by mixing human-induced pluripotent stem cells (hiPSC)-derived cardiomyocytes, fibroblasts, and/or other cell types such as endothelial cells to more closely mimic the structure and cellular complexity of native myocardium (Giacomelli et al. (2020); Saini et al. (2015)). However, a major obstacle in the development of EHTs is the inevitable trade-off between the need for miniaturization to increase throughput (i.e. several organoids in each well of a 96- or 384-well plate) and the biological complexity of the tissues (Cho et al. (2022)). The first EHTs were generated in small batches, using pre-designed molds, and having a millimetric size (Boudou et al. (2012); Turnbull et al. (2014); Legant et al. (2009); Mannhardt et al. (2016)). Since then, two different avenues of development have emerged. On the one hand, to optimize throughput, more straightforward, low complexity assays such as spheroids were developed to obtain smaller tissues (Beauchamp et al., 2015; Polonchuk et al., 2017). These multicellular constructs are valuable models to study drug responses and the human heart micro-environment but they lack the geometric intricacy of a native cardiac tissue. On the other hand, the search for higher tissue complexity and maturity led to the development of chambered organoids (MacQueen et al. (2018); Li et al. (2018); Lee et al. (2019)). These constructs require the use of complex techniques such as bioprinting or the use of bioreactors, and tissues are generated one at a time, making these approaches time-consuming and impractical on a large scale. Therefore, the next step in tissue engineering needs to be directed towards high-throughput physiologically-relevant assays that are simple and straightforward for the end-user. Scaling down the tissue size to increase throughput, while controlling their geometry represents a major engineering challenge. Moreover in actual designs, many EHTs are attached on static posts that can impact tissue formation, result in isometric contraction and can interfere with optical acquisitions when made with opaque material.

Here, we report the development of a novel assay which allows fast, reproducible and high-throughput generation of 3D cardiac rings in a 96-well plate using hiPSC-derived cardiomyocytes. The design allows the formation of multiple ring-shaped cardiac tissues in a well. The EHTs form around a central pillar made from an optically-transparent and deformable polymer that enables in situ monitoring of the tissue contraction with simultaneous measurements of force generation.

Results

Design and characterization of the molds

3D steel molds were designed to obtain a structure on which cells can be directed to ring-shaped cavities during sedimentation (Figure 1A). The mold fits in the well of a 96-well plate, and consists of 21 identical conical structures with a central hole (Figure 1A left). The molds obtained from the manufacturer were imaged by scanning electron microscopy and found to correspond to the intended design (Figure 1A, right). Casts of the mold were created in PEG to check the shape of the inner cavities of the mold, but the resulting structures were too soft to maintain their shape in air, and were transparent in water and therefore could not be imaged. More rigid polydimethylsiloxane (PDMS) was cast from the mold to verify the shape of the resulting structures (Figure 1B). These showed that the interior profile of the cavity was shaped in a reversed hourglass, as expected.

Description of the mold and seeding procedure. A. Micromachined stainless steel mold used to shape the gel – Design (left) and scanning electron microscopy image (right) of the mold – Scale bar: 500 µm. B. Molded PEG gel (left - scale bar 2 mm) and zoomed view of a pillar from its PDMS replica (middle-scale bar 100 µm). Design and size of the pillar (right). C. Timeline of the seeding procedure. D. Cardiac rings in a well one day after seeding. Images stitched with ImageJ plugin - Scale bar: 1 mm. E. Representative compaction of a ring with time after seeding (from day 0 to day 14), in brightfield - Scale bars: 200 µm.

Figure 1—video 1. Multiple rings beating at D14 - Brightfield imaging x4 magnification.

Generation and characterization of the gels

3D gel structures were obtained by adding a drop of PEG gel solution on top of the molds, inverting these onto a glass coverslip, polymerizing the gel and attaching the resulting sample on a 96-well plate. To determine the stiffness of the gels, a sample of the aqueous PEG solution was polymerized. A capillary was placed perpendicularly to the edge of the gel and negative pressure was applied to the gel through the capillary. The deformation of the gel into the capillary was measured and used to determine the Young’s modulus of the gel (Figure 1 supplement 1). The values obtained (11.4± 0.5 kPa) are within the physiological range for human cardiac tissue (Gershlak et al. (2013)).

Generation of ring-shaped cardiac tissues

The efficiency of the differentiation of iPSCs into cardiomyocytes was in average of 96.9 ± 0.8 % (Figure 1 Supplement 2).Cardiomyocytes derived from iPSCs (iPSC-CMs) with fibroblasts were seeded on the gels. After centrifugation, the cells sedimented into the ring-shaped cavities (Figure 1D), and compacted around the pillars to rapidly form ring-shaped cardiac constructs (Figure 1D-E). Initial tests with cardiomyocytes resulted in cell cluster formation and malformation of the rings (Figure 2A). To stabilize the tissues, iPSC-CMs were mixed with fibroblasts and different ratios of hiPSC-CMs to fibroblasts were tested in order to obtain the highest number of full stable cardiac rings with time. Three ratios of hiPSC-CMs:fibroblasts were tested (Figure 2B) based on previous reports (Tiburcy et al. (2017); Saini et al. (2015); Giacomelli et al. (2020); Beauchamp et al. (2020)). The 3:1 hiPSC-CMs:fibroblasts ratio was determined to be optimal as it allowed the generation of the highest number of tissues stable in time, and we could not see any difference in the contractile parameters of the tissues with different fibroblast ratios (see Figure 2 Supplement 1). Therefore, the 3:1 hiPSC-CMs:fibroblasts ratio was used for all of the following experiments. Tissues started beating within 24 hours after seeding (Video 1). The rings compacted in time as shown in Figure 1E. In these optimization experiments, we obtained an average of 10.75±0.48 tissues per well which were stable over time for at least 20 days with the 3:1 hiPSC-CMs:fibroblasts ratio.

Composition and structure of the cardiac rings. A.Example brightfield images of cardiomyocytes seeded with or without fibroblasts one day after seeding - Scale bar : 200 µm. B.Number of full cardiac rings per well in time according to the hiPSC-CMs:fibroblasts ratio they contain (4:1, 3:1 or 2:1). C.Confocal imaging of immunostained tissues at different heights: cardiac ring (top panel), and basal layer (bottom panel) of a tissue at 40X magnification. Vimentin, stained in green, corresponds to fibroblasts, troponin T, in red, is specific to cardiomyocytes, and DAPI is in blue. Scale bars: 100 µm. D.Picture of the immunostained contractile fibers at 63X magnification. Vimentin (green), Troponin T (red) and DAPI (blue) – Scale bar: 50µm. E.3D reconstruction of a ring. x, y and z scale bars : 100 µm.

Figure 2—video 1. 3D reconstruction of a ring

Structure and organization of the engineered cardiac tissues

The 3D organization of the tissues was assessed by 3D immunofluorescence imaging 14 days after seeding. The fibroblasts are concentrated at the base of the ring in contact with the glass (Figure 2C – bottom panel), whereas the cardiomyocytes form a compact ring above the fibroblasts around the central pillar (Figure 2C – top panel). Troponin T staining evidenced the formation of striated elongated fibers (Figure 2D), which are typical of cardiac tissue. The obtained cardiac tissues are 75.5±1.8 µm high and their shape is toric, as shown in the 3D reconstruction (Figure 2E and video 1). To investigate the kinetics of the fibroblasts-cardiomyocytes segregation, the same immunostainings were performed respectively 1 and 7 days after seeding (see Figure 2 Supplement 3). It appears that 1 day after seeding the fibroblasts are not yet attached, although the cardiac fiber has already started to be formed. Seven days after seeding, fibroblasts are fully spread and attached, and the contractile ring is formed and well-aligned.

Contractility analysis

Tissues started contracting less than 24 hours after seeding, which was visible through the deformation of their central pillar (Supplemental video 1). Each tissue was recorded in brightfield imaging, at a 10X magnification with a high-speed camera (Supplemental video 1). An in-house Matlab code was developed to quantify the contraction and relaxation phases of the tissues by monitoring the area of the central pillar over time (Figure 3A). The beat rate of the tissues first decreased between day 1 and day 3 and increased again to reach 0.58±0.030 Hz on day 14 (Figure 3B). The contraction stress developed by the tissues increased from day 1 until day 7 (Figure 3C). At day 7, the contractility parameters stabilized with an average stress of 1.4±0.1 mN/mm2 (Figure 3C.). The corresponding developed strain εA was of 24.84±0.92 % (Figure 3 supplement 1). Consistently, the relaxation and contraction speeds also increased during the first 7 days, then reaching a plateau (Figure 3D and E). Since long-term culture of iPSC-CMs is gaining interest we studied the contractile parameters of our rings 28 days after seeding in comparison to their contractile parameters at D14 (see Figure 3 Supplement 2). We found a slight increase for all the parameters, which is significant for the maximum contraction speed. Nevertheless, the data is much more variable and the number of tissues is lower (29 for D14 against 17 for D28). Therefore, we demonstrated that long-term culture of our tissues is possible, however not yet optimized. Hence, the following physiological and pharmacological tests have been done at D14.

Contractility analysis of the tissues. A.Principle of the in-house Matlab code used for contractility analysis: detection of the central pillar and monitoring of the evolution of its area in time, calculation and plot of the strain εA in time (ratio between the contraction amplitude in time and the maximum area of the central pillar) – Representative plots of the strain in time and its derivative in time, for a tissue at day 14. B. to E.Evolution of beating parameters through time after seeding at days 1, 3, 7, 10 and 14. The changes of all the parameters through time are significant (p<0.0001 - ANOVA for repeated measures −D1: n=57, D3: n=59 D7: n=47, D10: n=43, D14: n=36 tissues, from 3 differentiations). Beating parameters at each time point are compared to their value at day 1. B. Evolution of beat rate through time after seeding (***: p<0.002). C.Evolution of contraction stress through time after seeding (****: p<0.0001). D.Evolution of maximum contraction speed through time after seeding (****: p<0.0001). E.Evolution of maximum relaxation speed through time after seeding (****: p<0.0001).

Figure 3—video 1. Contractility of a ring at day 14.

Arrhythmia analysis

The regularity of the rings was first evaluated by plotting each period (time between two contraction peaks) as a function of the following one. We found that the Poincaré plot was fitted with a y=x line with R2 = 0.9633 (Figure 4 A) and with a narrow confidence interval, showing that the vast majority of the ntℎ periods are equal to the following ones, and therefore that the rings beat regularly without spontaneous extra beats. Also, following the study of Li et al. (2020) which shows the occurrence of re-entrant waves in the case of circular geometry of tissues, we performed live imaging of a voltage-sensitive fluorescent dye in our tissues and looked for re-entrant waves. We found that, as expected, the tissue depolarization induced a visible increase in fluorescence (see Figure 4 B.). All the pixels were synchronous and we could not detect any sparks or re-entrant waves (see the video 1). We then developed a Matlab script to analyze the average FluoVolt fluorescence signal together with the contraction over time, which was also obtained from the same videos. We found that the tissues beat regularly, with a depolarization occurring just before the contraction of the ring, causing a large increase in tissue fluorescence (from 0 to 6 ΔF ∕F0). We also observed that the repolarization signal is noisy, as it occurs during the time when the tissue deformation is greatest and is probably affected by tissue motion.

Study of arrhythmia in ring-shaped cardiac tissues. A. Poincaré plot for tissues at D14. n=24 tissues from 3 differentiations. B. Tissue with FluoVolt dye in its relaxed/hyperpolarized state and contracted/depolarized state. C. Representative plot of contraction strain (black) and fluorescence signal (green) of a tissue derived from video analysis of the FluoVolt dye.

Figure 4—video 1. FluoVolt live imaging of a ring at D15

Physiological testing on the engineered cardiac tissues

The response in contractility of the cardiac rings to an increase in the extracellular calcium concentration was assessed by sequential addition of calcium in the medium, starting from a concentration of 0.5 mM up to 3.5 mM (Figure 5A). The changes in the contraction stress and the contraction and relaxation speeds with calcium concentration were significant (p<0.0001 for each parameter, non significant for the beat rate). We observed a stabilization of the beat rate around 0.5Hz at 2mM and a positive inotropic response as the contraction amplitude increased by 600 % with a corresponding EC50 of 1.346 mM with a confidence interval (CI) of [1.159;1.535] mM.

Physiological and drug testing on the cardiac tissues. Effect of the concentration in extracellular calcium (A.), verapamil (B.), isoproterenol (C.) or dofetilide (D.) on tissues contractility: beat rate, contraction stress and the maximum contraction and relaxation speeds. These parameters are expressed as a ratio between their value for each concentration and the value at basal state ([Ca2+]=0.5 mM for calcium test and [Drug]=0M for drug testing). An ANOVA for repeated measures was carried out for each parameter. For each test, the value of each parameter at each concentration is compared to its value at the minimal concentration of the drug (respectively [Ca2+]=0.5mM, [Verapamil]=10−9 M, [Isoproterenol]=10−9 M and [Dofetilide]=10−9 M). E. Poincaré plot for 0 nM, 1 nM, 3 nM and 7 nM of dofetilide. For each concentration of each drug, more than 20 tissues from 3 different concentrations could be analyzed. Data is presented as mean ± SEM. *: p<0.05; **: p<0.01; ***: p<0.001; ****: p<0.0001.

Pharmacological testing on the engineered cardiac tissues

We next evaluated the dose-response to increasing concentrations of several cardiotropic drugs. The changes in the beating parameters with verapamil concentration were found to be significant (p=0.0053 for the beat rate and p<0.0001 for the other parameters). The addition of verapamil, an L-type calcium channel inhibitor, induced a negative inotropic effect until the complete stop of contraction at 10µM, as expected (Figure 5B). The IC50 was estimated to be 0.677 µM, CI=[4.278;12.81].107 M. The dose-response to the β-adrenergic receptor agonist, isoproterenol was tested. We observed a trend towards increase of the contraction amplitude, as well as a significant increase in the contraction and relaxation speeds (Figure 5C), without change in the beat rate. The changes in the contraction stress and the contraction and relaxation speeds with isoproterenol concentration are significant (p=0.0004 for the contraction stress; p<0.0001 for the contraction and relaxation speeds). Finally, dofetilide, which is an inhibitor of the IKr current, was also tested at different concentrations on the cardiac rings. This drug has been shown to be pro-arrhythmic and to induce arrhythmic events such as Torsades-de-Pointes in patients. The changes in the beat rate, the contraction stress and the contraction and relaxation speeds with dofetilide concentration were found to be significant (p<0.0001 for all the parameters). As shown in Figure 5D, dofetilide significantly reduced the contraction stress and speeds until the complete stop of the contraction of the cardiac tissues for a concentration of 10 nM. Arrhythmic events were noted after adding dofetilide at concentrations of 3 and 7 nM. Indeed, the Poincaré diagram for the different dofetilide concentrations (Figure 5E) shows an increase in the contraction-to-contraction intervals (equivalent to RR intervals) for some cells at 1 nM of dofetilide and an increasing number of erratic beats with increasing dofetilide concentrations.

Discussion

The combination of pluripotent stem cells technologies, advanced cardiomyocyte differentiation protocols and microfabrication methods have already resulted in the generation of engineered human cardiac muscle tissue-constructs, with different shapes and that can reproduce some cardiac functions (Seguret et al. (2021); Cho et al. (2022)). However, until now, approaching the complex geometry and structure of the human myocardium required complex techniques, which limit the number of tissues produced, thus hampering their application for high-throughput experiments. In this study, we aimed to address this challenge by developing a novel strategy to create ring-shaped miniature cardiac tissues around a central pillar that is made from an optically-transparent and deformable polymer. Our main findings are as follows : (1) this novel strategy enables the fast and reproducible formation of an important number of cardiac tissues (up to 21 per well in a 96-well format) with a limited number of hiPSC-cardiomyocytes (112,500 per well); (2) In this assay, fibroblasts are essential to generate and maintain the tissue structure, with a 3:1 hiPSC- CMs:fibroblast ratio providing with the best yield of tissue generation; (3) The self-organized ring-shaped cardiac tissues contract and deform the optically-transparent central pillar, which can be recorded and analyzed to estimate the contractile force developed by the tissue; (4) These ring-shaped tissues are compatible with live fluorescence imaging and allow the simultaneous measurement of contractile parameters and another readout provided by a fluorescent probe; (5) These EHTs display concordant responses to positive or negative inotropes, and to arrhythmogenic drugs, as required for drug testing applications.

The miniaturized cardiac constructs develop by self-organization of the hiPSC-derived cardiomyocytes and fibroblasts, which are guided by a specific design to acquire a ring shape. The resulting EHTs are in a convenient 96-well plate format and easy to use for a user familiar with cell biology techniques. The 96-well plate format is adaptable to many plate readers and automated platforms. In addition, the optical clarity of the gels and the absence of gel directly under the structures ensures optimal imaging conditions. We showed that the tissues can be fixed, stained by immunofluorescence and imaged in situ at high resolution. Moreover, our EHTs are also be compatible with live cell imaging experiments with fluorescent reporters, such as the FluoVolt voltage-sensitive dye, for which we presented a proof-of-concept. Indeed, we demonstrated that while measuring the depolarization of the cells by imaging the fluorescence over time, we could simultaneously monitor the contraction parameters in situ through the deformation of the pillar. This possibility to simultaneously record two parameters represents a significant asset, as this proof-of-concept can be extended to other fluorescent probes.

The current design of our EHTs consists of up to 21 rings in a single well, thus providing many replicates within a single experiment, even if some of the rings fail to form due to biological variability. This high throughput can compensate for potential intra-batch effects. Moreover, the possibility to seed several wells at the same time allows to easily study several differentiations, compensating for the variability which is inherent to the differentiation of hiPSCs. The predictable circular shape of the tissues makes them easy to recognize and track using software. Coupled with the known organization of the 21 rings for easy automated imaging, a single well can quickly generate data on many tissues that is quickly and automatically analyzed. The EHTs we described here are thus easy to use, image, manipulate and obtain readouts from.

We found that the circular geometry of these self-organized cardiac tissues induces a homogeneous distribution of the cardiomyocytes and consequently of the contraction forces around the pillar and in the tissue. Importantly, the presence of the central pillar not only allows resistance to be felt by the tissue, but the control of its stiffness facilitates the calculation of the force developed by the tissue. Our results showed that the tissues develop a fractional shortening (or contraction strain) of about 25%, which is close to the contraction of a human heart (Cheng et al. (2010)). The corresponding stress was 1.4 mN/mm2 which is in a similar range to other previously reported tissue-engineered cardiac muscle models with hiPSC-CMs in the literature (Seguret et al. (2021); Turnbull et al. (2014); Shadrin et al. (2017); Zhao et al. (2019); Ronaldson-Bouchard et al. (2018); Goldfracht et al. (2020); Li et al. (2020)). Some studies have reported higher force values (Ronaldson-Bouchard et al. (2018)) but in tissues that have a much large size than the tissues proposed here. The force generated by the adult human heart muscle, around 44 mM/mm2 for a tissue strip (Hasenfuss et al. (1991)), however remains significantly higher than the force developed by the currently available platforms. Few cardiac assays can precisely monitor the force exerted by the tissues. Indeed, cardiac cells seeded on micropillars (Rodriguez et al. (2014)) and gels containing tracking beads (Dou et al. (2021); Feyen et al. (2020)) can provide information on local (subcellular) force exertion. The force measurements do not result from the concerted effort of an organized tissue composed of aligned cardiomyocytes, as proposed by ring-shaped tissues. Tissues attached on two posts (Turnbull et al. (2014); Mannhardt et al. (2016); Abilez et al. (2018)) can also provide some information on the force generated by a tissue but the force exertion is not radial as it is in the native heart. Moreover, the presence of the posts can impact the contraction by affecting the force distribution in the tissue and creating potential edge effects in the vicinity of the posts, inducing the formation of v-necks (Abilez et al. (2018)). Lastly, some other ring-shaped cardiac constructs have to be transferred on posts or on a force transducer to measure their contraction force (Goldfracht et al. (2020); Tiburcy et al. (2017)). Therefore, the easy in situ monitoring of the force generated by our ring-shaped tissues with an image analysis routine represents a significant advantage, as less manipulation improves the degree of standardization. Finally, the stiffness of the central pillar can easily be adapted to model a pathological increase or decrease of the extracellular matrix stiffness. The EHTs we developed are composed of cardiomyocytes that rely on fibroblasts for their assembly and inter-cellular adhesion. Fibroblasts have been shown to play a prominent role in the heart as they are crucial in the constitution of the supporting extracellular matrix and contribute to cardiomyocyte electrical coupling, conduction system insulation, vascular maintenance, and stress-sensing (Ivey and Tallquist (2016); Baudino et al. (2006)). Other studies have highlighted the importance of a multi-cellular approach to generate EHTs (Giacomelli et al. (2020); Saini et al. (2015); Amano et al. (2016); Caspi et al. (2007)). In our system, the fibroblasts additionally play a role in providing a basement support under the cardiomyocytes, in contact with the glass support of the EHTs. Immunofluorescence imaging showed an organization of the cardiomyocytes parallel to each other, perpendicular to the direction of contraction, with an increased alignment of sarcomeres. Our EHTs mimic the composition and organization of the native myocardial tissue, even if only composed of cardiomyocytes and fibroblasts. Future experiments will determine whether other, less abundant, cell types of the native myocardium, such as endothelial or smooth muscle cells, can further improve heart tissue modeling.

The constructs showed a spontaneous beating in the day following seeding and this activity was maintained over several weeks thus allowing contractile measurements and pharmacological testing in the conditions of auto-pacing. Since Li et al. (2020) highlighted the occurrence of re-entrant waves in their circular cardiac tissues, we investigated the arrhythmogenicity of our tissues and performed live imaging of depolarization in the tissues to evaluate the presence of re-entrant waves. We were able to show that our cardiac tissues beat regularly, and we were able to correlate the depolarization with contraction over time. The next step would be to correct the fluorescence profile to account for tissue deformation in order to study tissue repolarization. This would require the development of a complex pixel-tracking code. The analysis of re-entrant waves is however hampered by a technical limitation. Indeed, it has been shown that the conduction velocity in iPSC-CMs is around 10-20 cm.s−1, compared to 40-60 cm.s−1 in adult cardiomyocytes Ahmed et al. (2020). Here we image our tissues at the maximum speed allowed by our camera, which is between 40 and 50 frames per second, which is faster than in Li et al. (2020). However, our constructs are smaller, with a perimeter of about 1.3 mm, and therefore, the required camera speed to detect the conduction in tissues should be at least of 120 fps if we consider a conduction speed of 15 cm.s−1. Hence we were not able to detect conduction in our tissues, nor re-entrant waves.

This first study is nevertheless a proof of concept for live imaging in miniature ring-shaped tissues, already allowing us to correlate the depolarization peaks with contraction peaks thus reflecting the excitation-contraction coupling efficiency in the tissues. Additional efforts, especially with a faster camera, will be needed to more accurately analyze tissue conduction and detect re-entrant waves. As a proof-of-concept, we also studied the effects of well-established negative and positive inotropes and showed the ability of our platform to reproduce the anticipated effects. We observed that our tissues presented a positive inotropic response to increasing extracellular calcium concentration with a corresponding EC50 which is in agreement with the values reported in the literature for engineered heart tissues made of hiPSC-CM, which are between 0.4 and 1.8mM for the EC50 (Feric and Radisic (2016); Schaaf et al. (2011); Streckfuss-Bömeke et al. (2013); Mannhardt et al. (2016); Turnbull et al. (2014); Goldfracht et al. (2020)). The negative inotropic response to increasing concentrations of verapamil of our constructs was also described in the literature and our IC50 value is also in the range of reported values in other EHTs: Turnbull et al. Turnbull et al. (2014) found IC50=0.61 µM and other papers range from 0.3 µM to 0.6 µM (Mannhardt et al. (2016); Thavandiran (2019)). The β-adrenergic stimulation of the tissues with increasing concentrations of isoproterenol-induced a trend towards increase in the contraction force and significantly increased contraction and relaxation speeds without change in the beat rate. This result may be surprising, as many articles report a positive chronotropic behavior for spontaneous hiPSC-CMs, while the inotropic response is less often highlighted (de Lange et al. (2021); Chen et al. (2015); Ronaldson-Bouchard et al. (2018)). Nevertheless the increase in contraction and relaxation speeds that we observed is consistent with data reported in the literature. Our hypothesis is that this increase in contraction and relaxation speeds induced by isoproterenol is translated, on average in our study, into an increase in contractile force rather than in an increase in contraction frequency. This may depend on the cell line used, and is very well illustrated in Mannhardt et al. (2020): of the 10 different cell lines tested in engineered heart tissues, all show an increase in contraction and relaxation speeds after isoproterenol administration, but this is translated either into an increase in contractile force (4 cell lines) or into a shortening of the beat (3 cell lines), and only 2 cell lines show an increase in both parameters. Indeed, since hiPSC-CMs are immature cardiac cells, it is rare to obtain a positive force-frequency relationship without any maturation medium or mechanical or electrical training. Last, applying increasing dofetilide concentrations on our constructs showed a decrease in the contraction amplitude and contraction and relaxation speeds. This is in line with the literature as Zhao et al. (2019) depicted a significant reduction of the action potential amplitude and an increase in action potential durations of the tissues with increasing dofetilide concentration, in line with Huo et al. (2019); Blinova et al. (2017). Above a concentration of 10nM, dofetilide shows cardiotoxicity in our tissues as tissues completely stop beating. By assessing the beating regularity by plotting each period as a function of the previous one for doses of 0nM to 7nM of dofetilide, we highlighted an increasing number of erratic beats with the increase in dofetilide concentration. Other papers have also underlined the pro-arrhythmic effect of dofetilide on cardiomyocytes derived from iPSCs (Goldfracht et al. (2019); Patel et al. (2019); Altrocchi et al. (2020); Blinova et al. (2017)) and this drug has been shown to induce atrial fibrillation, torsade-de-pointes and heart failure in patients (Abraham et al., 2015; Shantsila et al., 2007). Therefore, we demonstrated that our platform recapitulated the expected cardiac responses to several inotropes and arrhythmogenic drugs.

While representing a significant improvement, our platform presents some limitations. First, the size of the rings is very small compared to native tissue (400 µm). This size limitation is necessary for the purposes of high throughput screening, but for applications requiring larger tissues, wider rings can be considered in the future. Second, all measurements were performed under spontaneous pacing as a specific technique should be developed to allow external pacing of the tissues in the 96-well format. This includes electrical field stimulation or an optogenetics system with light stimulation. The precise and regular organization of the rings in the well makes this design very amenable to the introduction of microelectrodes for pacing, which could also allow the measurement of the action potentials of individual micro-tissues. Nevertheless, our main results show that pharmacological responses, including the detection of the pro-arrhythmogenic effects of dofetilide, were correctly measured under spontaneous beating conditions. Third, the miniaturized format of our platform creates some limitations and challenges regarding some experiments, including tissue stretching. Therefore, the length-tension relationships (Franck-Starling mechanism) were not determined in our tissues.

Overall the novel platform we describe here is a versatile, easy-to-use and high-throughput tool to generate cardiac rings in a reproducible way and analyze them in situ. This platform will be useful in several cardiac research fields, including disease modeling and pharmacological testing.

Methods and Materials

Mold design and characterization

Micromachined stainless steel molds were obtained from high resolution CNC machining according to our CAD designs. The designs consisted of a mold that would be slightly smaller than the footprint of a well of a 96-well plate, composed of 21 structures (see Figure 1A). To check the shape of the interior cavities of the mold, a cast of the mold was created out of polydimethylsiloxane (PDMS) (Sylgard 184, Samaro), according to manufacturer’s instructions.

3D hydrogel substrate preparation

A solution of 5% w/w photopolymerizable polyethylene glycol (PEG) was prepared in distilled water. Circular coverslips (16 mm, Paul Marienfeld GmbH, Germany) were silanized to ensure adequate bonding between the PEG and glass. A drop of 10-15 μL of gel was placed on the stainless steel mold, which was then placed in contact with a silanized coverslip. The gel solution was briefly exposed to light of the appropriate wavelength to polymerize the hydrogel. The resulting 3D structured hydrogel, bound to the coverslip, was unmolded and immediately placed in distilled water. The hydrogel was rinsed at least three times with distilled water and incubated in water overnight. For cell experiments, the coverslip coated in a 3D hydrogel was briefly dried and attached to a commercial black polystyrene 96-well plate with an adhesive backing (Grace Bio-labs, USA). The resulting well was filled with distilled water. Prior cell seeding, the plate was sterilized by UV irradiation for at least 20 minutes.

Young’s modulus measurements of the molds

The PEG-based hydrogel solutions were prepared and 100µL of solution was deposited into a ring-shaped mold with an inner diameter of 1cm positioned on a 16mm glass coverslip. The solution was polymerized using 365nm UV light for 1 minute at a power of 70 mW/cm2, after which the mold was removed. The coverslip containing the gel was then transferred to the microscope (Leica camera AG, Germany). A hollow glass capillary with an inner diameter 0.75mm (World Precision Instruments, USA) was attached to a platform movable in the xy-plane which was used to position the capillary right next to the edge of the circular polymerized hydrogel. Once the capillary and the hydrogel made contact, a negative pressure was applied, using the Cobalt autonomous pressure pump (Elvesys, France), creating a small convexity in the surface of the hydrogel. Images of this convexity were taken (5X magnification) for a pressure ranging from −100mbar to −200mbar, with 10 mbar increments (Figure 1A.). Two measurements were done on two separate gels, resulting in 4 measurements per gel with 10 datapoints each. The images (and thus the size of the convexities) were analyzed using ImageJ (version 1.53c). The Young’s modulus was calculated with the formula according to literature (Gandin et al. (2021)).

Cardiomyocyte differentiation and culture

The SKiPSC-31.3 hiPSC cell line, from a healthy control subject, has been used here and was previously reported (Galende et al. (2010)). Induced pluripotent stem cells were seeded on Matrigel and cultivated in mTESR Plus medium (Stemcell Technologies). At 80 % confluency, cells were passaged with ReleSR reagent (StemCell Technologies). When hiPSCs reached a 90 % confluency, differentiation was carried out according to a protocol adapted from Garg et al. (2018). Briefly, on day 0 of differentiation, mTESR Plus was changed to RPMI-1640 (Life Technologies) supplemented with B27 without insulin (RPMI-B27 minus insulin) and 6 µM CHIR-99021 (Abcam). After 48 hours, the medium was changed to RPMI-B27 minus insulin and then to RPMI-B27 minus insulin with 5µM IWR-1 (Sigma) for 48 hours. On day 5, the medium was replaced by RPMI-B27 minus insulin, and switched to RPMI-B27 with insulin on day 7. At day 11 a glucose starvation was carried out by replacing the medium with RPMI 1640 without glucose (Life Technologies), supplemented by B27 with insulin for 3 days. Cells were then dissociated with 0.05% trypsin (Life Technologies) and replated in a 12-well plate at a density of 0.3 106 cells/cm2 in RPMI-B27 with insulin. The following day, the medium was switched back to RPMI 1640 without glucose (Life Technologies), supplemented by B27 with insulin for 3 days again. These two rounds of glucose starvation have been shown to increase dramatically the percentage of cardiomyocytes obtained Sharma et al. (2015). After day 18, the cells were cultured in RPMI-B27 with insulin and the medium was changed every two days.

Cardiomyocyte characterization

The efficiency of the differentiation was assessed by cardiac troponin T flow cytometry at D21. On day 21 of differentiation, hiPSC-CMs were dissociated by enzymatic digestion (Miltenyi Multi Tissue dissociation kit 3) and stained with the Zombie NIR™ Fixable Viability Kit (BioLegend). Then, collected cell pellets were fixed and permeabilized using Inside Stain kit (Miltenyi Biotech, 130-090-477) at room temperature for 10 minutes. Cells were incubated with either APC anti-cardiac troponin T (CTNT) antibody (Miltenyi Biotech; 130-106-689 1:100) or APC isotype control (Miltenyi Biotech, 130-104-615 1:100) for 10 min at room temperature. Cells were analyzed using the BD Biosciences FACS LSR Fortessa X-20 instrument with at least 30,000 cells. Results were processed using FlowJo v10 (FlowJo, LLC).

Human fibroblasts culture

Commercially available normal adult human dermal fibroblasts (NHDF-Ad Lonza CC-2511, Lot 545147) were used up to passage 6 to generate the tissues. They were cultured in T75 flasks, in DMEM high glucose, supplemented with 10% FBS, 1% NEAA and 1% penicillin-streptomycin.

Tissues generation and culture

The wells of the assay were first rinsed twice with PBS and incubated at 37°C in DMEM high glucose, 2.3 mM CaCl2, 10%FBS, 0.1% penicillin-streptomycin. To generate the tissues, cardiomyocytes were dissociated by enzymatic digestion (Miltenyi Multi Tissue dissociation kit 3) on day 22 after differentiation. Fibroblasts were dissociated with TrypLE Express Enzyme (12605010 Thermofischer). Three ratios of hiPSC-CMs:fibroblasts were tested: 4:1, 3:1 and 2:1. For the following experiments the tissues are generated with a ratio of 3:1 hiPSC-CMs:fibroblasts as this ratio allowed the best preservation of tissues in time. A total of 150k cells were seeded in each well. The required numbers of fibroblasts and cardiomyocytes were then mixed and resuspended in 200µL per well of DMEM high glucose, 2.3 mM CaCl2, 10%FBS, 0.1% penicillin-streptomycin. The plate was then incubated at 37°C for 5 minutes and centrifuged 3 times in “short” mode to ensure that the cells fall into the circular molds. The plate was then kept at 37°C and the medium was changed every two days.

Tissues recording and analysis

The cardiac rings were recorded for 10 seconds in brightfield imaging with a high speed CCD camera (PL-D672MU, Pixelink) mounted on a microscope (Primovert, Zeiss), at a 10X magnification. The videos were then analyzed with a custom Matlab script. Briefly, the area of the central pillar is monitored in time and allows to recover the contraction amplitude in time Amplitude(t)= Max(Area)-Area(t) and therefore the contraction strain in time (in terms of area) εA calculated as εA(t) = (Max(Area) − Area(t))∕Max(Area) as well as the derivative of this strain in time. εA(t) is linked to the linear strain ε with the formula . As we measured the Young’s Modulus of the gel, we can recover the stress developed by the cardiac rings σ = E.ε(t). From these traces we also derive the contraction frequency by Fourier transform. It was considered that below a contraction of 0.5 %, the signal to noise ratio was not sufficient enough to detect the contraction of the tissue and the beating frequency was therefore set to 0 Hz. Finally, the maximum contraction and relaxation speeds of the tissue were also calculated respectively from the maxima and minima of the derivative of this strain in time.

Live imaging in tissues

Around 15 days after seeding, tissues were washed with a Tyrode solution (NaCl 14 mM, KCl 5 mM, 10 mM Hepes, Glucose 10 mM, MgCl2 1 mM, CaCl2 1.8 mM, pH 7.4) and incubated 30 minutes at 37°C, 5% Co_2 with the FluoVolt Loading solution (FluoVolt dye dilution 1:1000, Powerload solution dilution 1:100 in the Tyrode solution) as per the manufacturer’s recommendations (FluoVolt kit - Thermofisher). They were then washed with Tyrode and incubated in 200 µL Tyrode per well at 37°C, 5% CO2 to recover. Live fluorescence imaging of the Fluovolt dye was performed in situ with a Leica TSC SP8 confocal imaging system in resonant mode at 37°C incubation, with a 20X objective and a 256×256 pixels resolution. This allowed a frame rate of 43.96 frames per second.

Response to increasing extracellular calcium concentration and drug tests

Tissues response to different drugs and increasing calcium concentrations was assessed after day 14. For the increasing extracellular calcium concentrations, tissues were first changed to Tyrode solution 0.5 mM in calcium (NaCl 140 mM, KCl 5 mM, 10 mM Hepes, Glucose 10 mM, MgCl2 1 mM, CaCl2 0.5 mM, pH 7.4) and let to equilibrate for 30 minutes at 37°C, 5 % CO2. Calcium concentration was then increased by sequential addition of a Tyrode solution at 10mM in calcium. After each addition, the tissues were incubated for 5 minutes before recording. For drug tests, tissues were changed to Tyrode solution at 1.8mM in calcium (NaCl 14 mM, KCl 5 mM, 10 mM Hepes, Glucose 10 mM, MgCl2 1 mM, CaCl2 1.8 mM, pH 7.4) and let to equilibrate for 30 minutes at 37°C, 5% CO2. To evaluate the dose-response of the cardiac rings to the different drugs (Verapamil-sigma v0100000, isoproterenol - sigma 1351005, dofetilide - sigma PZ0016), the contractility was first recorded for 30 seconds in basal state and the drug was then added sequentially. Between each addition of drug, the tissues were incubated for 5 minutes at 37°C before being recorded for 30 seconds.

Immunofluorescence staining and confocal microscopy

Tissues were fixed at day 15 with 4% paraformaldehyde for 15 minutes and rinsed 3 times in PBS for 5 minutes. They were then blocked and permeabilized with blocking solution(BSA 2%, Triton 0.5%) at 4°C overnight. Primary antibodies Vimentin diluted at 1:250 (MA5-11883, ThermoFischer Scientific) and Troponin T diluted at 1:500 (ab45932, Abcam) were added the next day and incubated at 4°C overnight. Tissues were rinsed 3 times in PBS for 5 minutes and secondary antibodies (Alexa Fluor secondary antibodies, LifeTechnologies, at 1:1000) and DAPI diluted at 1:1000 were added and incubated at 4°C overnight. Immunostaining pictures were taken with a Leica TCS SP8 confocal system.

Statistical analysis

All numerical results are expressed as mean ± standard error mean (SEM) of three independent experiments. Differences between experimental groups were analyzed with the appropriate statistical tests, specified each time. P values < 0.05 were considered significant for all statistical tests. Statistical analyses were performed with GraphPad Prism software.

Supporting information